---
title: "UCell 官方完整数据 Workflow / Official Full Data Workflow"
lang: "zh-CN"
output:
  pdf_document:
    toc: true
    toc_depth: '3'
  html_document:
    toc: true
    toc_depth: 3
    theme: null
    css: styles/ucell_pdf.css
---

## 0. 目标 / Goal

本文件使用 UCell 官方 vignette 中的 `scRNAseq::ZilionisLungData()` 数据，走一遍不含 Seurat 的完整 UCell workflow。
This notebook uses the official UCell vignette data source, `scRNAseq::ZilionisLungData()`, and runs the full non-Seurat UCell workflow.

本轮覆盖 / Covered here:

- 官方 Zilionis lung cancer single-cell 数据下载与缓存。Download and cache the official Zilionis lung cancer single-cell data.
- 筛选 immune cells，并取前 5000 cells。Filter immune cells and use the first 5000 cells.
- Matrix workflow: `ScoreSignatures_UCell()`.
- 参数学习：`maxRank`、`w_neg`、`missing_genes`、`chunk.size`。Parameter exploration.
- rank 缓存：`StoreRankings_UCell()` + `precalc.ranks`。Precomputed rank workflow.
- `SingleCellExperiment` workflow。

不包含 / Not covered:

- Seurat workflow.
- `SmoothKNN()` 实跑。

## 1. Setup

```{r setup}
project_root <- "F:/WorkSpace/Causality+AI-multiGRN"
knitr::opts_knit$set(root.dir = project_root)
setwd(project_root)

.libPaths(c("F:/WorkSpace/Causality+AI-multiGRN/r-lib", .libPaths()))

library(UCell)
library(BiocParallel)
library(scRNAseq)
library(SingleCellExperiment)
library(SummarizedExperiment)
library(Matrix)
library(ggplot2)

set.seed(123)
BPPARAM <- SnowParam(workers = 1)

dir.create("data", showWarnings = FALSE)
dir.create("results", showWarnings = FALSE)

cache_file <- "data/ucell_zilionis_immune_5000.rds"
```

说明 / Notes:

- 本项目 Windows 环境使用 `SnowParam(workers = 1)`。This Windows project uses `SnowParam(workers = 1)`.
- 下载后的处理数据会保存到 `data/ucell_zilionis_immune_5000.rds`。Processed data will be cached as an RDS file.

## 2. Helper Functions

这些函数只负责整理表格和缓存数据，不改变 UCell 算法。
These helper functions only organize tables and cache data. They do not modify the UCell algorithm.

```{r helpers}
make_long_scores <- function(score_matrix) {
  data.frame(
    cell = rep(rownames(score_matrix), times = ncol(score_matrix)),
    signature = rep(colnames(score_matrix), each = nrow(score_matrix)),
    score = as.numeric(as.matrix(score_matrix)),
    row.names = NULL
  )
}

make_gene_overlap <- function(signatures, expr_mat) {
  rows <- lapply(names(signatures), function(signature_name) {
    genes <- signatures[[signature_name]]
    clean_genes <- sub("[+-]$", "", genes)
    data.frame(
      signature = signature_name,
      gene = genes,
      clean_gene = clean_genes,
      direction = ifelse(grepl("-$", genes), "negative", "positive"),
      present_in_matrix = clean_genes %in% rownames(expr_mat),
      row.names = NULL
    )
  })
  do.call(rbind, rows)
}
```

## 3. 下载或读取缓存数据 / Download or Read Cached Data

第一次运行会下载官方数据；之后会直接读取本地缓存。
The first run downloads the official data. Later runs read the local cache.

```{r load-or-cache-data}
if (file.exists(cache_file)) {
  message("Reading cached data: ", cache_file)
  cached <- readRDS(cache_file)
} else {
  message("Downloading ZilionisLungData() through scRNAseq...")
  lung <- scRNAseq::ZilionisLungData()

  immune <- lung$Used & lung$used_in_NSCLC_immune
  lung <- lung[, immune]
  lung <- lung[, seq_len(min(5000, ncol(lung)))]

  assay_name <- if ("counts" %in% assayNames(lung)) {
    "counts"
  } else {
    assayNames(lung)[1]
  }

  exp.mat <- Matrix::Matrix(assay(lung, assay_name), sparse = TRUE)
  colnames(exp.mat) <- paste0(colnames(exp.mat), seq_len(ncol(exp.mat)))

  cached <- list(
    exp.mat = exp.mat,
    colData = as.data.frame(colData(lung)),
    assay_name = assay_name,
    source = "scRNAseq::ZilionisLungData; immune cells; first 5000 cells"
  )

  saveRDS(cached, cache_file)
}

exp.mat <- cached$exp.mat

class(exp.mat)
dim(exp.mat)
cached$assay_name
```

检查 / Checks:

```{r data-checks}
stopifnot(inherits(exp.mat, "dgCMatrix"))
stopifnot(ncol(exp.mat) <= 5000)

head(rownames(exp.mat))
head(colnames(exp.mat))
summary(Matrix::colSums(exp.mat > 0))
```

解释 / Interpretation:

- `exp.mat` 是 genes x cells。`exp.mat` is genes x cells.
- 这是 UCell matrix workflow 的标准输入方向。This is the standard input orientation for UCell.
- `Matrix::colSums(exp.mat > 0)` 表示每个 cell 检测到的基因数。This shows detected genes per cell.

## 4. 官方 Signatures / Official Signatures

这些 signatures 来自 UCell 官方 vignette。
These signatures are from the official UCell vignette.

```{r official-signatures}
signatures <- list(
  Tcell = c("CD3D", "CD3E", "CD3G", "CD2", "TRAC"),
  Myeloid = c("CD14", "LYZ", "CSF1R", "FCER1G", "SPI1", "LCK-"),
  NK = c("KLRD1", "NCR1", "NKG7", "CD3D-", "CD3E-"),
  Plasma_cell = c("MZB1", "DERL3", "CD19-")
)

signatures
```

检查 signature 基因是否在矩阵里。
Check whether signature genes are present in the matrix.

```{r gene-overlap}
gene_overlap <- make_gene_overlap(signatures, exp.mat)
gene_overlap

write.csv(
  gene_overlap,
  "results/ucell_zilionis_gene_overlap.csv",
  row.names = FALSE
)
```

说明 / Notes:

- 末尾 `-` 表示负向 marker。A trailing `-` means a negative marker.
- 例如 `LCK-` 表示如果 LCK 高表达，会降低 Myeloid score。For example, `LCK-` penalizes the Myeloid score if LCK is high.

## 5. Matrix Workflow: `ScoreSignatures_UCell()`

这是最核心的官方调用。
This is the core official function call.

```{r score-matrix}
ucell_scores <- ScoreSignatures_UCell(
  exp.mat,
  features = signatures,
  BPPARAM = BPPARAM
)

head(ucell_scores)
dim(ucell_scores)
summary(ucell_scores)
```

结果方向 / Output orientation:

- 行是 cell。Rows are cells.
- 列是 signature。Columns are signatures.
- 列名自动添加 `_UCell`。Column names automatically get the `_UCell` suffix.

```{r save-scores}
write.csv(ucell_scores, "results/ucell_zilionis_scores.csv")
saveRDS(ucell_scores, "results/ucell_zilionis_scores.rds")

ucell_scores_long <- make_long_scores(ucell_scores)
write.csv(
  ucell_scores_long,
  "results/ucell_zilionis_scores_long.csv",
  row.names = FALSE
)

head(ucell_scores_long)
```

## 6. Score Distribution

```{r plot-distribution}
ggplot(ucell_scores_long, aes(x = signature, y = score, fill = signature)) +
  geom_violin(scale = "width", alpha = 0.75) +
  geom_boxplot(width = 0.12, outlier.size = 0.4, alpha = 0.85) +
  theme_bw(base_size = 12) +
  labs(
    title = "UCell score distributions on Zilionis immune cells",
    x = "Signature",
    y = "UCell score"
  ) +
  theme(legend.position = "none")
```

解释 / Interpretation:

- violin 显示每个 signature 在 5000 个 immune cells 中的分布。The violin plot shows score distributions across 5000 immune cells.
- boxplot 显示中位数和四分位数。The boxplot shows median and quartiles.

## 7. `maxRank` 参数比较 / Compare `maxRank`

`maxRank` 控制 UCell 排名截断。
`maxRank` controls the rank cutoff.

```{r compare-maxrank}
scores_800 <- ScoreSignatures_UCell(
  exp.mat,
  features = signatures,
  maxRank = 800,
  BPPARAM = BPPARAM
)

scores_1500 <- ScoreSignatures_UCell(
  exp.mat,
  features = signatures,
  maxRank = 1500,
  BPPARAM = BPPARAM
)

scores_3000 <- ScoreSignatures_UCell(
  exp.mat,
  features = signatures,
  maxRank = 3000,
  BPPARAM = BPPARAM
)

compare_maxrank <- data.frame(
  cell = rownames(scores_1500),
  Tcell_800 = as.numeric(scores_800[, "Tcell_UCell"]),
  Tcell_1500 = as.numeric(scores_1500[, "Tcell_UCell"]),
  Tcell_3000 = as.numeric(scores_3000[, "Tcell_UCell"]),
  row.names = NULL
)

head(compare_maxrank)
write.csv(
  compare_maxrank,
  "results/ucell_zilionis_compare_maxRank.csv",
  row.names = FALSE
)
```

```{r plot-maxrank}
ggplot(compare_maxrank, aes(x = Tcell_1500, y = Tcell_800)) +
  geom_point(size = 0.7, alpha = 0.45) +
  geom_abline(slope = 1, intercept = 0, color = "red") +
  theme_bw(base_size = 12) +
  labs(
    title = "Tcell score: maxRank 1500 vs 800",
    x = "maxRank = 1500",
    y = "maxRank = 800"
  )

cor(compare_maxrank$Tcell_1500, compare_maxrank$Tcell_800)
```

## 8. 正负基因与 `w_neg` / Positive-Negative Signatures and `w_neg`

这里使用官方 `Myeloid` signature，其中 `LCK-` 是负向 marker。
Here we use the official `Myeloid` signature, where `LCK-` is a negative marker.

```{r compare-wneg}
myeloid_signature <- list(
  Myeloid = c("CD14", "LYZ", "CSF1R", "FCER1G", "SPI1", "LCK-")
)

myeloid_w1 <- ScoreSignatures_UCell(
  exp.mat,
  features = myeloid_signature,
  w_neg = 1,
  BPPARAM = BPPARAM
)

myeloid_w05 <- ScoreSignatures_UCell(
  exp.mat,
  features = myeloid_signature,
  w_neg = 0.5,
  BPPARAM = BPPARAM
)

compare_wneg <- data.frame(
  cell = rownames(myeloid_w1),
  w_neg_1 = as.numeric(myeloid_w1[, "Myeloid_UCell"]),
  w_neg_0.5 = as.numeric(myeloid_w05[, "Myeloid_UCell"]),
  LCK_expr = as.numeric(exp.mat["LCK", ]),
  row.names = NULL
)

head(compare_wneg)
summary(compare_wneg$w_neg_0.5 - compare_wneg$w_neg_1)
```

```{r plot-wneg}
ggplot(compare_wneg, aes(x = w_neg_1, y = w_neg_0.5, color = LCK_expr)) +
  geom_point(size = 0.7, alpha = 0.55) +
  geom_abline(slope = 1, intercept = 0, color = "red") +
  theme_bw(base_size = 12) +
  labs(
    title = "Effect of w_neg on Myeloid score",
    x = "w_neg = 1",
    y = "w_neg = 0.5",
    color = "LCK expression"
  )
```

解释 / Interpretation:

- `w_neg = 1` 表示负向 marker 全权重惩罚。`w_neg = 1` uses full penalty for negative markers.
- `w_neg = 0.5` 表示负向惩罚减半。`w_neg = 0.5` halves the negative penalty.

## 9. 缺失基因处理 / Missing Genes

```{r missing-genes}
missing_signature <- list(
  Tcell_missing = c("CD3D", "CD3E", "CD2", "not_a_gene")
)

scores_impute <- ScoreSignatures_UCell(
  exp.mat,
  features = missing_signature,
  missing_genes = "impute",
  BPPARAM = BPPARAM
)

scores_skip <- ScoreSignatures_UCell(
  exp.mat,
  features = missing_signature,
  missing_genes = "skip",
  BPPARAM = BPPARAM
)

compare_missing <- data.frame(
  cell = rownames(scores_impute),
  impute = as.numeric(scores_impute[, "Tcell_missing_UCell"]),
  skip = as.numeric(scores_skip[, "Tcell_missing_UCell"]),
  row.names = NULL
)

head(compare_missing)
summary(compare_missing$skip - compare_missing$impute)
```

## 10. `chunk.size`

`chunk.size` 影响内存和速度，不应影响结果。
`chunk.size` affects memory and runtime, not the scores.

```{r chunk-size}
scores_chunk_100 <- ScoreSignatures_UCell(
  exp.mat,
  features = signatures,
  chunk.size = 100,
  BPPARAM = BPPARAM
)

scores_chunk_500 <- ScoreSignatures_UCell(
  exp.mat,
  features = signatures,
  chunk.size = 500,
  BPPARAM = BPPARAM
)

all.equal(scores_chunk_100, scores_chunk_500)
```

## 11. 预计算 Rank / Precompute Ranks

如果要反复测试多个 signatures，先保存 ranks 会更高效。
If you repeatedly test many signatures, precomputing ranks is more efficient.

```{r store-ranks}
ranks <- StoreRankings_UCell(
  exp.mat,
  maxRank = 1500,
  BPPARAM = BPPARAM
)

class(ranks)
dim(ranks)
ranks[1:5, 1:5]

saveRDS(ranks, "results/ucell_zilionis_ranks.rds")
```

用预计算 ranks 重新打分。
Rescore using precomputed ranks.

```{r score-from-ranks}
scores_from_ranks <- ScoreSignatures_UCell(
  features = signatures,
  precalc.ranks = ranks,
  BPPARAM = BPPARAM
)

head(scores_from_ranks)
write.csv(
  scores_from_ranks,
  "results/ucell_zilionis_scores_from_ranks.csv"
)
```

比较直接打分与 ranks 打分。
Compare direct scoring and precomputed-rank scoring.

```{r compare-direct-ranks}
max_abs_delta <- max(abs(as.matrix(ucell_scores) - as.matrix(scores_from_ranks)))
max_abs_delta
```

新增 signatures，不需要重新计算 ranks。
Score new signatures without recomputing ranks.

```{r new-signatures}
new.signatures <- list(
  Lymphoid = c("LCK"),
  CD3_module = c("CD3D", "CD3E", "CD3G")
)

scores_new <- ScoreSignatures_UCell(
  features = new.signatures,
  precalc.ranks = ranks,
  BPPARAM = BPPARAM
)

head(scores_new)
```

## 12. SingleCellExperiment Workflow

UCell 可以直接对 `SingleCellExperiment` 对象打分。
UCell can directly score a `SingleCellExperiment` object.

```{r sce-workflow}
sce <- SingleCellExperiment(list(counts = exp.mat))

sce <- ScoreSignatures_UCell(
  sce,
  features = signatures,
  assay = "counts",
  BPPARAM = BPPARAM
)

altExpNames(sce)
stopifnot("UCell" %in% altExpNames(sce))

sce_scores <- assay(altExp(sce, "UCell"), "UCell")
sce_scores[1:4, 1:5]

saveRDS(sce, "results/ucell_zilionis_sce_with_scores.rds")
```

解释 / Interpretation:

- matrix 输入会直接返回 score matrix。Matrix input returns a score matrix.
- SCE 输入会返回原 SCE，并把结果放到 `altExp(sce, "UCell")`。SCE input returns the SCE object with scores in `altExp(sce, "UCell")`.

## 13. Workflow Summary

```{r workflow-summary}
summary_table <- data.frame(
  metric = c(
    "n_genes",
    "n_cells",
    "n_signatures",
    "max_abs_delta_direct_vs_precalc_ranks",
    "sce_has_ucell_altExp"
  ),
  value = c(
    nrow(exp.mat),
    ncol(exp.mat),
    length(signatures),
    max_abs_delta,
    "UCell" %in% altExpNames(sce)
  )
)

summary_table
write.csv(
  summary_table,
  "results/ucell_zilionis_workflow_summary.csv",
  row.names = FALSE
)
```

## 14. 关键结论 / Key Takeaways

- UCell 官方 matrix 输入是 genes x cells。Official matrix input is genes x cells.
- `ScoreSignatures_UCell()` 是核心打分函数。`ScoreSignatures_UCell()` is the core scoring function.
- 官方 Zilionis workflow 可先筛选 immune cells 并取 5000 cells 学习流程。The official Zilionis workflow can be learned on 5000 immune cells first.
- `StoreRankings_UCell()` 适合反复测试多个 signatures。`StoreRankings_UCell()` is useful when testing many signatures repeatedly.
- `SingleCellExperiment` 输入会把分数写入 `altExp(sce, "UCell")`。For SCE input, scores are stored in `altExp(sce, "UCell")`.

## 15. Session Info

```{r session-info}
sessionInfo()
```
